Conceptual and methodological problems with bias detection and avoidance in natural language processing

### 2021-06-21 thesis/01-intro.Rmd
Adds review remarks

Natural language processing (NLP) is a subfield of computer science that processes and analyzes language in text and speech with the use of modern programming methods. It has practical applications in everyday life as it concerns tasks such as email filters, smart assistants, search results, language translations, text analytics and so on. Models used to accomplish these tasks need a lot of data to learn from. This data originates from human activities and historical recordings such as texts, messages or speeches. It turns out that in the learning process these models can learn implicit biases that reflect harmful stereotypical thinking still present in modern societies. There are many different types of models in NLP depending on a task that they are supposed to solve. However, all of them need as an input words represented by means of numbers and this is accomplished with word embeddings. The word embeddings are usually assigned the values based on the context in which the words appear. This means that the input data can have an enormous influence on the outcome. The biases seem to have their primary source in the way the words are assigned numerical values. One can find methods that aim at identifying and measuring hidden biases and/or try to remove them by for example modifying the word embeddings directly.

One can find various definitions trying to capture what bias and fairness actually are. With the choice of the definition, implications into the real-life applications may change as well. @Mehrabi2019Survey mark out that there exist different types of biases such as historical bias, representation bias or measurement bias (the list is long). This indicates how complex the issue of bias is. Without the proper understanding and awareness of the problem, people are prone to unconsciously sustain the bias. @Mehrabi2019Survey also distinguish different types of discrimination, some of them will be briefly described. By protected attributes we mean those qualities, traits or characteristics that one should not discriminate against. Direct discrimination occurs when protected attributes of individuals explicitly result in non-favorable outcomes toward them. Such discrimination takes place for example when one uses the information about someone's religious beliefs to evaluate whether they are competent for the work they apply for. In contrast, in indirect discrimination individuals appear to be treated equally, but anyway they end up being treated unjustly due to the hidden effects of biases towards their protected attributes. An example of such situation is when in a job advert there is a requirement of 10 years experience instead of the list of a specific type of experience and knowledge. This job advert can discriminate indirectly against young people who can have the required skills, but not yet so much work experience. Systemic discrimination takes place when policies, customs or behaviors that result from certain culture or organizational structure lead to discrimination against some groups of people. Finally, statistical discrimination consists in relying on group statistics to judge person belonging to a given group. The topic of discrimination is entangled with another concept; fairness. It is essential to grasp some concepts of fairness to take them into consideration while designing implementation of some machine learning model. In @Mehrabi2019Survey one may notice that depending on the context and application different definitions may be applied.

There is a considerable amount of literature available on the topic of bias detection and mitigation in NLP models. @Bolukbasi2016Man focus on gender biases that may be observable while investigating the representation of job occupations and gender in terms of their assigned numerical
values. The authors apply cosine similarity measurement to investigate the phenomenon where (the vectors corresponding to) words related to jobs that are stereotypically associated with a given gender are in fact in the model situated closer to this gender. 
They also use analogy tasks to evaluate if bias is present in a word embedding model. They check analogies by comparing pairs of word vectors. For example they search for the word complementing the puzzle: man is to doctor as woman is to ...? First they subtract the word "man" from the word "woman"^[This is pointwise subtraction of vectors, but for the simplicity we will often talk about words while really meaning vectors. What is meant should be clear from the context.] and then 
they search for the ranked list of other words pairs that have similar vector difference. They also include in the formula a threshold to ensure that the resulting pairs could not be randomly picked.

However, as @Nissim2019Fair points out, there are some limitations to this approach. In practice, most of analogies implementations do not return any input words. This means that it does not make sense to expect the algorithm to return the same profession for both woman and man. Therefore, this method seems to be limited in terms of bias detection. Other problems are related for example to the choice of pairs and words that are used to detect the presence of discrimination, as it is often subjective and without proper justification. Additionally, the choice of the parameter used in @Bolukbasi2016Man formula to ensure that word pairs are not picked by random, is also not justified and changing it drastically 
influences the results.

@@

@Manzini2019blackToCriminal propose a novel way of using cosine similarity to verify if the assumed resemblance between protected words and harmful attributes exists. They investigate an approach that enables them to measure the bias for a class (such as gender, religion or race) and express the final result with a single metric. Their approach will be analyzed in details in the next chapter.

@Caliskan2017Semantics refer to Implicit Association Test (Greenwald et al., 1998) that measures the strength of associations between concepts or stereotypes by measuring human reaction time for special tasks. Humans naturally exhibit some biases which do not always cause social concern. One can imagine the intuitive associations between for example insects and flowers, and the feelings of pleasantness or unpleasantness. In general, people would rather associate flowers with feeling pleasant than insects, and this preference in a sense is a bias or
prejudice in some direction. However, this type of preference does not cause an uproar and is a rather morally neutral case. This example is rather used to show that the methodology used in IAT makes sense. @Caliskan2017Semantics were able to obtain similar results to the ones that measured reaction time with the use of WEAT metric. Unfortunately, further studies discover other biases and prejudices that directly influence the quality of other people's lives and therefore they should be taken care of.

The most common bias detection methods in the literature focus on comparing the similarity between words from protected groups and those that are considered to be stereotypical or harmful in some way. One can find in this group methods such as euclidean distance or cosine similarity (which is equivalent to dot product if the vectors are normalized). There are also other ways to detect the effects of biases. One can investigate the model performance on certain downstream tasks and validate if the model uses the gender or race information to complete the task. The currently used methods employing cosine similarity, often aggregate the similarity values often aggregated in a way that may lead to hasty conclusions. The averaging of values and the lack of attention to uncertainty involved in such aggregation may lead to the incomplete picture of the situation.

>>>>

Natural language processing (NLP) is a subfield of computer science that processes and analyzes language in text and speech with the use of modern programming methods. It has practical applications in everyday life as it concerns tasks such as email filters, smart assistants, search results, language translations, text analytics and so on. Models used to accomplish these tasks need a lot of data to learn from. This data originates from human activities and historical recordings such as texts, messages or speeches. It turns out that in the learning process these models can learn implicit biases that reflect harmful stereotypical thinking still present in modern societies. There are many different types of models in NLP depending on a task that they are supposed to solve. However, all of them need as an input words represented by means of numbers and this is accomplished with word embeddings. Word embeddings are usually assigned the values based on the context in which the words appear. This means that the input data can have an enormous influence on the outcome. One can find methods that aim at identifying and measuring hidden biases and/or try to remove them by, for example, modifying the word embeddings directly.

One can find various definitions trying to capture what bias and fairness actually are. With the choice of the definition, implications for the real-life applications may change as well. @Mehrabi2019Survey mark out that there exist different types of biases such as historical bias, representation bias or measurement bias (the list is long). This indicates how complex the issue of bias is. Without the proper understanding and awareness of the problem, people are prone to unconsciously sustain the bias. @Mehrabi2019Survey also distinguish different types of discrimination, some of them will be briefly described. By protected attributes we mean those qualities, traits or characteristics that one should not discriminate against. Direct discrimination occurs when protected attributes of individuals explicitly result in non-favorable outcomes toward them. Such discrimination takes place for example when one uses the information about someone's religious beliefs to evaluate whether they are competent for the work they apply for. In contrast, in indirect discrimination individuals appear to be treated equally, but anyway they end up being treated unjustly due to the hidden effects of biases towards their protected attributes. An example of such situation is when in a job advert there is a requirement of 10 years experience instead of the list of a specific type of experience and knowledge. This job advert can discriminate indirectly against young people who can have the required skills, but not yet so much work experience. Systemic discrimination takes place when policies, customs or behaviors that result from certain culture or organizational structure lead to discrimination against some groups of people. Finally, statistical discrimination consists in relying on group statistics to judge a person belonging to a given group. The topic of discrimination is entangled with another concept; fairness. It is essential to grasp some concepts of fairness to take them into consideration while designing implementation of some machine learning model. In @Mehrabi2019Survey one may notice that depending on the context and application different definitions may be applied.

There is a considerable amount of literature available on the topic of bias detection and mitigation in NLP models. @Bolukbasi2016Man focus on gender biases that may be observable in the representation of job occupations and gender in terms of numerical
values. The authors apply cosine similarity measurement to investigate the phenomenon where (the vectors corresponding to) words related to jobs that are stereotypically associated with a given gender are in fact in the model situated closer to this gender. 
They also use analogy tasks to evaluate if a bias is present in a word embedding model. They check analogies by comparing pairs of word vectors. For example they search for the word complementing the puzzle: man is to doctor as woman is to ...? First they subtract the word "man" from the word "woman"^[This is pointwise subtraction of vectors, but for the simplicity we will often talk about words while really meaning vectors. What is meant should be clear from the context.] and then 
they search for the ranked list of other words pairs that have similar vector difference. They also include in the formula a threshold to ensure that the resulting pairs could not be randomly picked.

However, as @Nissim2019Fair points out, there are some limitations to this approach. In practice, most of analogies implementations do not return any input words. This means that it does not make sense to expect the algorithm to return the same profession for both woman and man. Therefore, this method of bias detection seems limited. Other problems are related for example to the choice of pairs and words that are used to detect the presence of discrimination, as it is often subjective and without proper justification. Additionally, the choice of the parameter used in @Bolukbasi2016Man formula to ensure that word pairs are not picked by random, is also not justified and changing it drastically 
influences the results.

@@

@Manzini2019blackToCriminal propose a novel way of using cosine similarity to verify if the assumed resemblance between protected words and harmful attributes exists. They develop an approach that enables them to measure the bias for a class (such as gender, religion or race) and express the final result with a single metric. Their approach will be analyzed in details in the next chapter.

@Caliskan2017Semantics refer to Implicit Association Test (Greenwald et al., 1998) that measures the strength of associations between concepts or stereotypes by measuring human reaction time for special tasks. Humans naturally exhibit some biases which do not always cause social concern. One can imagine the intuitive associations between for example insects and flowers, and the feelings of pleasantness or unpleasantness. In general, people would rather associate flowers than insects with feeling pleasant, and this preference in a sense is a bias or
prejudice in some direction. However, this type of preference does not cause an uproar and is a rather morally neutral case. This example is rather used to show that the methodology used in IAT makes sense. @Caliskan2017Semantics were able to obtain similar results to the ones that measured reaction time with the use of WEAT metric. Unfortunately, further studies discover other biases and prejudices that directly influence the quality of other people's lives and therefore they should be taken care of.

The most common bias detection methods in the literature focus on comparing the similarity between words from protected groups and those that are considered to be stereotypical or harmful in some way. One can find in this group methods such as Euclidean distance or cosine similarity (which is equivalent to dot product if the vectors are normalized). There are also other ways to detect the effects of biases. One can investigate the model performance on certain downstream tasks and validate if the model uses the gender or race information to complete the task. Unfortunately, the currently used methods employing cosine similarity, often aggregate the similarity values in a way that may lead to hasty conclusions. The averaging of values and the lack of attention to uncertainty involved in such aggregation may lead to an incomplete picture of the situation.

### 2021-06-21 thesis/02-cosine.Rmd
Adds review remarks

The numbers occurring in such representations are not random. They are learned in a process that uses a very shallow neural network. There are various types of techniques used for learning the vectors representations. One of the most straightforward ones is a skip-gram model.
Given a word the model tries to predict its neighboring words from the sentence. The mathematics behind the process relies on the idea that the prediction concerns the conditional probability of the adjacent words. The algorithm tries to minimize the loss function, which penalizes the system for discrepancy with actual co-occurrence frequencies in the corpus. One can choose various parameters of the model, such as the window size that determines how many surrounding words the model should predict. After preparing such a fitted model, one takes only the learned weights from a neural network, and uses them as vectors in a word embeddings representation.

@@

\begin{align} \tag{Sim}
\mathsf{cosineSimilarity}(A,B) & = \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert}
\end{align}
<!-- add information about inner product role -->
Cosine similarity is considered a proper tool for this operation as its result has a clear connection to geometry and at least for a low number of dimensions may be easily interpreted. Using this scale, one can compare vector similarities in a fairly clear manner. When the vectors are aligned perpendicularly to each other, their similarity equals 0 (which is the same as the cosine of 90 degrees). This tells us that the similarity between the vectors is small. As the angle between vectors decreases, cosine similarity approaches one, which stands for the greatest similarity. It is also possible to obtain negative cosine similarity, if the value approaches -1 it means that the words are contrary to each other.

@@

One should note that cosine distance is not exactly a distance measure, as it does not meet triangle inequality requirements. The triangle inequality formula says that for any triangle, the sum of the lengths of any two sides must be greater than or equal to the length of the remaining side. As shown in [discussion in stats.stackexchange.com](https://stats.stackexchange.com/questions/198080/proving-that-cosine-distance-function-defined-by-cosine-similarity-between-two-u) in the case of cosine distance it would have to fulfill this equation 1+cos-sim(𝐴,𝐶) < cos-sim(𝐴,𝐵) + cos-sim(𝐵,𝐶). If one chooses specific unit vectors it is easy to demonstrate that the triangle inequality is not preserved.

@@

@Bolukbasi2016Man define similarity between words as the outcome inner product of their normalized vectors. 
They focus on examining what the geometry of word embedding is in regard to "he" and "she" words. In other words, whether the similarity between those concepts and other words reflects expected gender stereotypes. They test this hypothesis by investigating whether there is a connection between word embeddings representing certain professions and words referring to gender. They also evaluate whether automatically produced analogies between words reflect the stereotypes as well.

@@

@Manzini2019blackToCriminal present a different approach towards finding similarities between classes of words. The authors claim that texts available online are full of direct or indirect human stereotypes. As a result, word embeddings are prone to learn and maybe amplify those biases and propagate them further into AI models that are used for various applications. Cosine distance is used in the article as a measure to first argue for the existence of multi-class bias and then to show how through bias mitigation techniques the bias may be decreased.

@@

In the final step there is an evaluation of the results. The cosine distances are calculated again but this time using debiased vocabulary. After taking the mean of all distances one final value is obtained and then it is compared with the average value from the beginning. If the cosine distances are on average greater than before then it leads the authors to the conclusion that improvement has been achieved. As the cosine distance increases, it is assumed that the association between protected and stereotypical words decreases.

@@

The authors use the mean of mean average cosine similarities to measure multi-class similarity between protected word and harmful stereotypes. They average the results until they obtain one final value to represent the mean cosine distance between protected word from a given class and the attributes of that class. If one takes closer look at the individual values that are taken for the calculations it turns out that there is a bunch of outliers and surprisingly dissimilar words. We approach this issue by providing the tables and visualizations of individual cosine distances to make sure that we get the proper insight into the data.

@@

With such a method the uncertainty is not included which makes it even more difficult to give reasonable interpretations of the results. We propose the use of Bayesian method to obtain some understanding of the influence the uncertainty has on the interpretation of final results.

\textbf{4. No word class distinction}

In the original paper, words from all three religions were compared against all of the stereotypes, which means that there was no distinction between cases in which the stereotype is associated with a given religion, as opposed to the situation in which it is associated with another one. Not all of the stereotypical words have to be considered as harmful for all of the religions. In our analysis, we distinguish between stereotypes associated with a given group, stereotypes associated with different groups, and control groups: neutral words and stereotypes-free human predicates.

@@

MAC stand for mean average cosine similarity, although in reality the the table contains mean distances. What may attract attention is the fact that the value of cosine distance in "Biased" category is already quite high even before debiasing. High cosine distance indicates low cosine similarity between values. One could think that average cosine similarity equal to approximately 0.141 is not significant enough to consider it as bias. However the authors aim to mitigate "biases" in vectors with such great distance to make it even larger. Methodologically the question is, on what basis is this small similarity still considered as a proof of the presence of bias, and whether these small changes are meaningful. This is in general the problem of scale and the lack of universal intervals. In contrast, proper intervals decide whether given cosine similarity is high enough to consider the words to be more similar than if we chose them at random. We will use highest posterior density intervals.

>>>>

The numbers occurring in such representations are not random. They are learned in a process that uses a very shallow neural network. There are various types of techniques used for learning the vectors representations. One of the most straightforward ones is a skip-gram model.
Given a word, the model tries to predict its neighboring words from the sentence. The mathematics behind the process relies on the idea that the prediction concerns the conditional probability of the adjacent words. The algorithm tries to minimize the loss function, which penalizes the system for discrepancy with actual co-occurrence frequencies in the corpus. One can choose various parameters of the model, such as the window size that determines how many surrounding words the model should predict. After preparing such a fitted model, one takes only the learned weights from a neural network, and uses them as vectors in a word embeddings representation.

@@

\begin{align} \tag{Sim}
\mathsf{cosineSimilarity}(A,B) & = \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert}
\end{align}
<!-- add information about inner product role -->
Cosine similarity is considered a proper tool for this operation as its result has a clear connection to geometry and at least for a low number of dimensions may be easily interpreted. Using this scale, one can compare vector similarities in a fairly clear manner. When the vectors are aligned perpendicularly to each other, their similarity equals 0 (which is the same as the cosine of 90 degrees). This tells us that the similarity between the vectors is small. As the angle between vectors decreases, cosine similarity approaches one, which stands for the greatest similarity. It is also possible to obtain negative cosine similarity. If the value approaches -1, this intuitively means that the words are contrary to each other.

@@

One should note that cosine distance is not exactly a distance measure, as it does not meet triangle inequality requirements. The triangle inequality formula says that for any triangle, the sum of the lengths of any two sides must be greater than or equal to the length of the remaining side. As shown in [discussion in stats.stackexchange.com](https://stats.stackexchange.com/questions/198080/proving-that-cosine-distance-function-defined-by-cosine-similarity-between-two-u)^[https://stats.stackexchange.com/questions/198080/proving-that-cosine-distance-function-defined-by-cosine-similarity-between-two-u] in the case of cosine distance it would have to fulfill this equation 1+cos-sim(𝐴,𝐶) < cos-sim(𝐴,𝐵) + cos-sim(𝐵,𝐶). If one chooses specific unit vectors it is easy to demonstrate that the triangle inequality is not preserved.

@@

@Bolukbasi2016Man define similarity between words as the outcome inner product of their normalized vectors. 
They focus on examining what the geometry of a word embedding is in regard to "he" and "she" words. In other words, whether the similarity between those concepts and other words reflects expected gender stereotypes. They test this hypothesis by investigating whether there is a connection between word embeddings representing certain professions and words referring to gender. They also evaluate whether automatically produced analogies between words reflect the stereotypes as well.

@@

@Manzini2019blackToCriminal present a different approach towards finding similarities between classes of words. Cosine distance is used in the article as a measure to first argue for the existence of multi-class bias and then to show how through bias mitigation techniques the bias may be decreased.

@@

In the final step they evaluate the results. The cosine distances are calculated again but this time using the debiased vocabulary. After taking the mean of all distances one final value is obtained and then it is compared with the average value from the beginning. If the cosine distances are on average greater than before then it leads the authors to the conclusion that improvement has been achieved. As the cosine distance increases, it is assumed that the association between protected and stereotypical words decreases.

@@

The authors use the mean of mean average cosine similarities to measure multi-class similarity between protected word and harmful stereotypes. They average the results until they obtain one final value to represent the mean cosine distance between protected word from a given class and the attributes of that class. If one takes closer look at the individual values that are taken for the calculations it turns out that there is a bunch of outliers and surprisingly dissimilar words. We approach this issue by providing the tables and visualizations of individual cosine distances to make sure that we obtain a proper insight into the data.

@@

With such a method the uncertainty involved is not really considered which makes it even more difficult to give reasonable interpretations of the results. We propose the use of Bayesian method to obtain some understanding of the influence the uncertainty has on the interpretation of final results.

\textbf{4. No word class distinction and no control group}

In the original paper, words from all three religions were compared against all of the stereotypes, which means that there was no distinction between cases in which the stereotype is associated with a given religion, as opposed to the situation in which it is associated with another one. Not all of the stereotypical words have to be considered as harmful for all of the religions. One should investigate the religions separately as some of them may have stronger harmful associations that others. One should also include control groups to have a way of comparing the stereotypical results with neutral or human-like words. Later in the text we will explain in details reasons for introducing control groups. In our analysis, we distinguish between stereotypes associated with a given group, stereotypes associated with different groups, and control groups: neutral words and stereotypes-free human predicates.

@@

MAC stand for mean average cosine similarity, although in reality the the table contains mean distances. What may attract attention is the fact that the value of cosine distance in "Biased" category is already quite high even before debiasing. High cosine distance indicates low cosine similarity between values. One could think that average cosine similarity equal to approximately 0.141 is not significant enough to consider it as bias. However the authors aim to mitigate "biases" in vectors with such great distance to make it even larger. Methodologically the question is, on what basis is this small similarity still considered as a proof of the presence of bias, and whether these small changes are meaningful. This is in general the problem of scale and the lack of universal intervals. In contrast, statistical intervals will help us decide whether a given cosine similarity is high enough to consider the words to be more similar than if we chose them at random. We will use highest posterior density intervals, in line with Bayesian methodology.

### 2021-06-21 thesis/03-bayesian.Rmd
Adds review remarks

We will use the choice of protected words and stereotypical predicates used in REF. This is a decent point of departure,  not only we want to compare our method to that of REF, but also because this data format is fairly general (as contrasted, say, with a set up for binary stereotypes). Note also that the method we develop here  can fairly easily be  run  for different stereotypization patterns.  Let's start  with  explaining the method and its deployment using a dataset obtained for the religion-related protected words.

Let's load, clean a bit and inspect the head of the religion dataset we prepared.  In order to obtain this  dataset, we  calculated the cosine distance between each protected word and each word from both the bias-related attribute groups, which were used in the original study, and to neutral and human control attributes which we added  as control groups.  For instance, for religion, the bias-related predicates (coming from the original study in REF) include muslim bias attributes, jew bias attributes, christian bias attributes (see a list in the APPENDIX).

@@

The first impression is that while  there is a  shift for  associated words towards smaller cosine distances as compared to the neutral words, slightly surprisingly a slightly weaker shift in the same direction is visible for attributes associated with different stereotypes. Moreover,  the empirical distributions overlap to alarge extent and the  means grouped by connection type do not seem too far from each other. In fact, as there is a lot of  variety in the cosine distances (as we will soon see), we need to gauge the uncertainty involved, and  to look more carefully at individual protected words to get a better idea of how the cosine distance distribution changes for different attribute groups and different protected classes.   Now, let's add the human attributes to the picture:

@@

The only reason we need it is the evaluation of connection as a predictor. Does including it in o the model improve the situation? To investigate this, let's now build a model according to the following specification:

@@

Let's turn then to data analysis that takes a look at protected words separately. This time for each combination of a protected word and a connection status we will have a separate mean cosine distance estimate, each coming with its own highest posterior density interval. This means we will use indices that are result from all such combinations (and then we will split them up in the model precis to build visualisation, feel free to look at the `visualiseStats.R` script for details).

@@

We first encountered the GoogleNews word embedding model in @Bolukbasi2016Man. They use it for their calculations as written in [Tolga Code](https://github.com/tolga-b/debiaswe). We decided to compare the results obtained with the use of Reddit L2 model and the ones that we got by GoogleNews model. The details of the model can be found here [Google News model](https://code.google.com/archive/p/word2vec/). One of the main differences between the models is that Reddit word embeggings have 50 dimensions and GoogleNews word embeddings have 300 dimensions. As the dimension is increasing, the vector can capture much more information but this is not always the case. According to some researchers the amount of dimensions should be chosen by taking into account the type of corpus and its features. In our case both Reddit and GoogleNews models are already trained and ready to use so we do not analyse further the choice of their hyperparameters.

@@

First, let's look at coefficient estimated for the whole datasets, as compared to their estimation prior to debiasing:

@@

One may observe very slight changes in the estimated coeﬀicients. It seems that in religion dataset the changes are the most significant as the mean moves towards zero which stands for no similarity between words.

@@

It is worth paying attention to the gender debiased dataset. The change is really minor, still zero out of HPDI range. The `different` attributes still have high similarity value. It seems that the distances for gender protected words are still clustered together which suggests that this method is not able to catch the complex bias presence. One could conclude that co-occurrence of certain terms does not always mean direct connection in terms of terms associations. The fact that some of stereotypically associated with female attributes still have high similarity with male protected words suggest that other techniques should be applied to detect and remove the bias.

@@

In @Manzini2019blackToCriminal it is mentioned that debiasing with this method that is based on bias component removal may be insufficient. The reason is that words containing similar biases may still cluster together even after debiasing.

>>>>

We will use the choice of protected words and stereotypical predicates used in @Manzini2019blackToCriminal. This is a decent point of departure, as not only we want to compare our method to that of @Manzini2019blackToCriminal, but also because this data format is fairly general (as contrasted, say, with a set up for binary stereotypes). Note also that the method we develop here  can fairly easily be  run  for different stereotypization patterns.  Let's start  with  explaining the method and its deployment using a dataset obtained for the religion-related protected words.

Let's load, clean a bit and inspect the head of the religion dataset we prepared.  In order to obtain this  dataset, we  calculated the cosine distance between each protected word and each word from both the bias-related attribute groups, which were used in the original study, and to neutral and human control attributes which we added  as control groups.  For instance, for religion, the bias-related predicates (coming from the original study in @Manzini2019blackToCriminal) include muslim bias attributes, jew bias attributes, christian bias attributes (see a list in the APPENDIX).

@@

The first impression is that while  there is a  shift for  associated words towards smaller cosine distances as compared to the neutral words, slightly surprisingly a slightly weaker shift in the same direction is visible for attributes associated with different stereotypes. Moreover,  the empirical distributions overlap to a large extent and the  means grouped by connection type do not seem too far from each other. In fact, as there is a lot of  variety in the cosine distances (as we will soon see), we need to gauge the uncertainty involved, and  to look more carefully at individual protected words to get a better idea of how the cosine distance distribution changes for different attribute groups and different protected classes.   Now, let's add the human attributes to the picture:

@@

\normalsize
The only reason we need it is the evaluation of connection as a predictor. Does including it in o the model improve the situation? To investigate this, let's now build a model according to the following specification:

@@

Let's turn then to data analysis that takes a look at protected words separately. This time, for each combination of a protected word and a connection status we will have a separate mean cosine distance estimate, each coming with its own highest posterior density interval. This means we will use indices that are result from all such combinations (and then we will split them up in the model precis to build visualisation, feel free to look at the `visualiseStats.R` script for details).

@@

We first encountered the GoogleNews word embedding model in @Bolukbasi2016Man. They use it for their calculations as written in [Tolga Bolukbasi code](https://github.com/tolga-b/debiaswe)^[https://github.com/tolga-b/debiaswe]. We decided to compare the results obtained with the use of Reddit L2 model and the ones that we got by GoogleNews model. The details of the model can be found here [Google News model](https://code.google.com/archive/p/word2vec/). One of the main differences between the models is that Reddit word embeggings have 50 dimensions and GoogleNews word embeddings have 300 dimensions. As the dimension increases, the vectors can capture much more information, but this is not always the case. According to some researchers, the amount of dimensions should be chosen by taking into account the type of corpus and its features. In our case both Reddit and GoogleNews models are already trained and ready to use so we do not analyse further the choice of their hyperparameters.

@@

First, let's look at coefficient estimated for the whole datasets, as compared to their estimation prior to debiasing. One may observe very slight changes in the estimated coeﬀicients. It seems that in religion dataset the changes are the most significant as the mean moves towards zero which stands for no similarity between words.

@@

@@

It is worth paying attention to the gender debiased dataset. The change is really minor, still with zero out of the HPDI range. The `different` attributes still have high similarity value. It seems that the distances for gender protected words are still clustered together which suggests that this method is unable to catch the complex bias presence. One could conclude that co-occurrence of certain terms does not always mean direct connection in terms of terms associations. The fact that some of stereotypically associated with female attributes still have high similarity with male protected words suggest that other techniques should be applied to detect and remove the bias.

@@
